---
title: "08-annotate-DMR"
author: "Yaamini Venkataraman"
date: "2024-08-07"
output: html_document
---

In this script I will examine overlaps between DMR and DEG for specific contrasts, then visualize DMR patterns. As a reminder, there were no DMR found for Scorton Creek populations, so this script will focus only on New Bedford Harbor populations when examining DMR-DEG overlaps.

# Set up R Markdown document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/yaaminivenkataraman/Documents/killifish-hypoxia-RRBS/output/08-annotate-DMR/new-genome/") #Set root directory
```

```{r}
getwd()
```

#Install packages

```{r}
# if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
# if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer')
# if ("ComplexHeatmap" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install('ComplexHeatmap')
# if ("patchwork" %in% rownames(installed.packages()) == 'FALSE') install.packages('patchwork')
# if ("ggExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('ggExtra')
require(tidyverse)
require(RColorBrewer)
require(ComplexHeatmap)
require(patchwork)
require(ggExtra)
```

```{r}
sessionInfo()
```

# Import and format data

## Methylation data

```{r}
NBH20vOCDMR <- read.delim("20_OC_N_DMR.closestGene.bed", header = FALSE, 
                          col.names = c("chr", "DMR.start", "DMR.end", "DMR", "direction", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strands", "V13", "V14", "distance")) %>%
  separate_wider_delim(., cols = "V14", delim = ";Dbxref", names = c("gene.name", "trash")) %>%
  mutate(., gene.name = gsub(x = gene.name, pattern = "ID=gene-", replacement = "")) %>%
  dplyr::select(chr, DMR.start, DMR.end, DMR, direction, gene.name, gene.start, gene.end, strands, distance) %>%
  left_join(., (read.delim("../../06-DMR/new-genome/missing_1/20_OC_N/20_OC_N_qval.0.05.bedgraph", header = FALSE,
                           col.names = c("chr", "DMR.start", "DMR.end", "meth.diff"))), by = c("chr", "DMR.start", "DMR.end"))
#Import DMR data and add column names. Separate column with gene name by a specific delimiter, then remove characters before gene name. Retain only necessary columns. Join by imported dataframe to get methylation difference at each DMR (treatment - outside control).
head(NBH20vOCDMR) #Confirm formatting
```

```{r}
NBH5vOCDMR <- read.delim("5_OC_N_DMR.closestGene.bed", header = FALSE, 
                         col.names = c("chr", "DMR.start", "DMR.end", "DMR", "direction", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strands", "V13", "V14", "distance")) %>%
  separate_wider_delim(., cols = "V14", delim = ";Dbxref", names = c("gene.name", "trash")) %>%
  mutate(., gene.name = gsub(x = gene.name, pattern = "ID=gene-", replacement = "")) %>%
  dplyr::select(chr, DMR.start, DMR.end, DMR, direction, gene.name, gene.start, gene.end, strands, distance) %>%
  left_join(., (read.delim("../../06-DMR/new-genome/missing_1/5_OC_N/5_OC_N_qval.0.05.bedgraph", header = FALSE,
                           col.names = c("chr", "DMR.start", "DMR.end", "meth.diff"))), by = c("chr", "DMR.start", "DMR.end"))
#Import DMR data and add column names. Separate column with gene name by a specific delimiter, then remove characters before gene name. Retain only necessary columns. Join by imported dataframe to get methylation difference at each DMR (treatment - outside control). #Reorganize manually with slice
head(NBH5vOCDMR) #Confirm formatting
```

```{r}
NBH20vOCmetilene <- read.delim("../../05-analysis/new-genome/summarize/missing_1/20_OC_N/N_metilene_NO_OC.txt", header = TRUE) #Import metilene data used to identify DMR. This dataset has the methylation value for each CpG in the DMR.
head(NBH20vOCmetilene) #Confirm import
```

```{r}
NBH20vOCpositions <- data.frame() #Create empty dataframe
for (i in 1:nrow(NBH20vOCDMR)) {
  temp <- NBH20vOCmetilene %>%
    filter(., pos >= NBH20vOCDMR$DMR.start[i] & pos <= NBH20vOCDMR$DMR.end[i]) %>%
    mutate(., DMR = NBH20vOCDMR$DMR[i])
  NBH20vOCpositions <- rbind(NBH20vOCpositions, temp)
} #Use a for loop to filter CpG data for each DMR. Filter positions based on start and end positions of each DMR. Add a column for DMR identity. Save as a temporary dataframe. Use rbind to append information to empty dataframe
colnames(NBH20vOCpositions) <- c(colnames(NBH20vOCmetilene), "DMR") #Assign column names based on previous dataframe and add name for the DMR column
head(NBH20vOCpositions) #Confirm formatting
```

```{r}
NBH20vOCavgMeth <- NBH20vOCpositions %>%
  group_by(., DMR) %>%
  dplyr::select(., -c(chr, pos)) %>%
  summarize_all(funs(mean(., na.rm = TRUE))) %>%
  slice(., 1, 3:10, 2) %>%
  column_to_rownames(var = "DMR") #Group by DMR. Remove chromosome and position columns. Average methylation level across all CpG positions in each DMR for each sample. Reorganize manually with slice. Change DMR column to rownames
head(NBH20vOCavgMeth) #Confirm formatting
```

```{r}
NBH5vOCmetilene <- read.delim("../../05-analysis/new-genome/summarize/missing_1/5_OC_N/N_metilene_HY_OC.txt", header = TRUE) #Import metilene data used to identify DMR. This dataset has the methylation value for each CpG in the DMR.
head(NBH5vOCmetilene) #Confirm import
```

```{r}
NBH5vOCpositions <- data.frame() #Create empty dataframe
for (i in 1:nrow(NBH5vOCDMR)) {
  temp <- NBH5vOCmetilene %>%
    filter(., pos >= NBH5vOCDMR$DMR.start[i] & pos <= NBH5vOCDMR$DMR.end[i]) %>%
    mutate(., DMR = NBH5vOCDMR$DMR[i])
  NBH5vOCpositions <- rbind(NBH5vOCpositions, temp)
} #Use a for loop to filter CpG data for each DMR. Filter positions based on start and end positions of each DMR. Add a column for DMR identity. Save as a temporary dataframe. Use rbind to append information to empty dataframe
colnames(NBH5vOCpositions) <- c(colnames(NBH5vOCmetilene), "DMR") #Assign column names based on previous dataframe and add name for the DMR column
head(NBH5vOCpositions) #Confirm formatting
```

```{r}
NBH5vOCavgMeth <- NBH5vOCpositions %>%
  group_by(., DMR) %>%
  dplyr::select(., -c(chr, pos)) %>%
  summarize_all(funs(mean(., na.rm = TRUE))) %>%
  slice(., 1, 12, 23, 34, 45, 56:59, 2:11, 13:22, 24:33, 35:44, 46:55) %>%
  column_to_rownames(var = "DMR") #Group by DMR. Remove chromosome and position columns. Average methylation level across all CpG positions in each DMR for each sample. Reorganize manually with slice. Change DMR column to rownames
head(NBH5vOCavgMeth) #Confirm formatting
```

## Gene expression data

```{r}
NBH20vOCDEG <- read.csv("../../../data/NEW-n.NBHvso.NBH.csv") #Import DEG data
colnames(NBH20vOCDEG)[1] <- "gene.name"
head(NBH20vOCDEG)
```

```{r}
NBH5vOCDEG <- read.csv("../../../data/NEW-h.NBHvso.NBH.csv") #Import DEG data
colnames(NBH5vOCDEG)[1] <- "gene.name"
head(NBH5vOCDEG)
```

```{r}
NBH20geneExp <- rbind((read.delim("../../../data/RNA-Seq-renamed/htseq-20-N1.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "20-N1_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-20-N2.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "20-N2_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-20-N4.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "20-N4_exp"))) %>%
  pivot_wider(., names_from = "sample", values_from = "CPM") #Import gene expression data and rename columns. Add a new column for sample ID. Use rbind to combine dataframes. Pivot wider.
head(NBH20geneExp) #Confirm import and formatting
```

```{r}
NBH5geneExp <- rbind((read.delim("../../../data/RNA-Seq-renamed/htseq-5-N1.txt", header = FALSE, 
                                 col.names = c("gene.name", "CPM")) %>%
                        mutate(., sample = "5-N1_exp")),
                     (read.delim("../../../data/RNA-Seq-renamed/htseq-5-N2.txt", header = FALSE, 
                                 col.names = c("gene.name", "CPM")) %>%
                        mutate(., sample = "5-N2_exp")),
                     (read.delim("../../../data/RNA-Seq-renamed/htseq-5-N3.txt", header = FALSE, 
                                 col.names = c("gene.name", "CPM")) %>%
                        mutate(., sample = "5-N3_exp"))) %>%
  pivot_wider(., names_from = "sample", values_from = "CPM") #Import gene expression data and rename columns. Add a new column for sample ID. Use rbind to combine dataframes. Pivot wider.
head(NBH5geneExp) #Confirm import and formatting
```

```{r}
NBHOCgeneExp <- rbind((read.delim("../../../data/RNA-Seq-renamed/htseq-OC-N1.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "OC-N1_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-OC-N2.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "OC-N2_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-OC-N3.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "OC-N3_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-OC-N4.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "OC-N4_exp")),
                      (read.delim("../../../data/RNA-Seq-renamed/htseq-OC-N5.txt", header = FALSE, 
                                  col.names = c("gene.name", "CPM")) %>%
                         mutate(., sample = "OC-N5_exp"))) %>%
  pivot_wider(., names_from = "sample", values_from = "CPM") #Import gene expression data and rename columns. Add a new column for sample ID. Use rbind to combine dataframes. Pivot wider.
head(NBHOCgeneExp) #Confirm import and formatting
```

# DMR figures

## Heatmaps

I will make heatmaps showing methylation level of DMR across samples for the normoxia vs. outside control and hypoxia vs. outside control comparisons.

```{bash}
mkdir figures
```

```{r}
heatmapColors <- rev(brewer.pal("RdBu", n = 11))
```

### Normoxia vs. outside control

```{r}
NBH20vOCoxygenTreatment <- data.frame(samples = colnames(NBH20vOCavgMeth),
                                      Treatment = c(rep("Mild", times = 3),
                                                    rep("Control", times = 4))) %>%
  column_to_rownames(var = "samples") #Create dataframe with sample ID and treatment information. Convert sample to rownames.
head(NBH20vOCoxygenTreatment) #Column annotation information
```

```{r}
NBH20vOCoxygenTreatmentColors <- list(Treatment = c("Mild" = "grey80",
                                                    "Control" = "grey20")) #List of colors for column annotations
NBH20vOCoxygenTreatmentColors #Column annotation colors
```

```{r}
#pdf("figures/20_OC_N-DMR-heatmap.pdf", width = 11, height = 8.5)

pheatmap(as.matrix(NBH20vOCavgMeth*100), cluster_row = TRUE, show_rownames = TRUE, cluster_cols = FALSE, show_colnames = FALSE, color = heatmapColors, annotation_col = NBH20vOCoxygenTreatment, annotation_colors = NBH20vOCoxygenTreatmentColors, annotation_names_col = FALSE, name = "Methylation (%)")

#dev.off()
```

### Hypoxia vs. outside control

```{r}
NBH5vOCoxygenTreatment <- data.frame(samples = colnames(NBH5vOCavgMeth),
                                     Treatment = c(rep("Severe", times = 3),
                                                   rep("Control", times = 4))) %>%
  column_to_rownames(var = "samples") #Create dataframe with sample ID and treatment information. Convert sample to rownames.
head(NBH5vOCoxygenTreatment) #Column annotation information
```

```{r}
NBH5vOCoxygenTreatmentColors <- list(Treatment = c("Severe" = "grey80",
                                                   "Control" = "grey5")) #List of colors for column annotations
NBH5vOCoxygenTreatmentColors #Column annotation colors
```

```{r}
#pdf("figures/5_OC_N-DMR-heatmap.pdf", width = 11, height = 8.5)

pheatmap(as.matrix(NBH5vOCavgMeth*100), cluster_row = TRUE, show_rownames = TRUE, cluster_cols = FALSE, show_colnames = FALSE, color = heatmapColors, annotation_col = NBH5vOCoxygenTreatment, annotation_colors = NBH5vOCoxygenTreatmentColors, annotation_names_col = FALSE, name = "Methylation (%)")

#dev.off()
```

## Hyper- vs. hypomethylation

I want to make a barplot showing the level of hyper- or hypomethylation of a DMR.

### Normoxia vs. outside control

```{r}
NBH20vOCDMR %>%
  mutate(order = as.numeric(gsub(x = DMR, pattern = "DMR_", replacement = ""))) %>%
  arrange(order) %>%
  ggplot(aes(y = factor(order), x = meth.diff)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(name = "Mean Methylation Difference",
                     breaks = seq(-1.0, 1.0, 0.25),
                     limits = c(-1,1)) +
  scale_y_discrete(name = "DMR") +
  theme_classic(base_size = 15)
ggsave("figures/20_OC_N-methDiff.pdf", height = 11, width = 8.5)
```

### Hypoxia vs. outside control

There are three DMR between hypoxia and the outside control that have two annotations:DMR_22, DMR_26, and DMR_36 have one annotation. I need to ensure I'm only plotting one value for each DMR.

```{r}
NBH5vOCDMR %>%
  dplyr::select(., -c(strands, gene.name, gene.start, gene.end, distance)) %>%
  unique(.) %>%
  mutate(order = as.numeric(gsub(x = DMR, pattern = "DMR_", replacement = ""))) %>%
  arrange(order) %>%
  ggplot(aes(y = factor(order), x = meth.diff)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(name = "Mean Methylation Difference",
                     breaks = seq(-1.0, 1.0, 0.25),
                     limits = c(-1,1)) +
  scale_y_discrete(name = "DMR",
                   breaks = seq(1, 59, 2)) +
  theme_classic(base_size = 15)
ggsave("figures/5_OC_N-methDiff.pdf", height = 11, width = 8.5)
```

### Multipanel plot

```{r}
NBH20vOCmethDiffPlot <- NBH20vOCDMR %>%
  mutate(order = as.numeric(gsub(x = DMR, pattern = "DMR_", replacement = ""))) %>%
  arrange(order) %>%
  ggplot(aes(y = factor(order), x = meth.diff)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(-1.0, 1.0, 0.25),
                     limits = c(-1,1)) +
  labs(x = "Mean Methylation Difference",
       y = "DMR",
       title = "A. Mild Hypoxia vs. Control") +
  theme_classic(base_size = 15)
```

```{r}
NBH5vOCmethDiffPlot <- NBH5vOCDMR %>%
  dplyr::select(., -c(strands, gene.name, gene.start, gene.end, distance)) %>%
  unique(.) %>%
  mutate(order = as.numeric(gsub(x = DMR, pattern = "DMR_", replacement = ""))) %>%
  arrange(order) %>%
  ggplot(aes(y = factor(order), x = meth.diff)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(-1.0, 1.0, 0.25),
                     limits = c(-1,1)) +
  scale_y_discrete(breaks = seq(1, 59, 2)) +
  labs(x = "Mean Methylation Difference",
       y = "DMR",
       title = "B. Severe Hypoxia vs. Control") +
  theme_classic(base_size = 15)
```

```{r}
NBH20vOCmethDiffPlot | NBH5vOCmethDiffPlot #Use patchwork to create a multipanel plot
ggsave("figures/multipanel-methDiff.pdf", height = 8.5, width = 11)
```

# Overlaps between DEG and DMR

```{r}
NBH20vOCDMRDEG <- inner_join(NBH20vOCDMR, NBH20vOCDEG, by = "gene.name") %>%
  mutate(contrast = "20_OC_N") #Use inner join to identify DMR that overlap with DEG.
NBH20vOCDMRDEG
```

```{r}
NBH5vOCDMRDEG <- inner_join(NBH5vOCDMR, NBH5vOCDEG, by = "gene.name") %>%
  mutate(contrast = "5_OC_N") #Use inner join to identify DMR that overlap with DEG.
NBH5vOCDMRDEG
```

```{r}
allNBHDMRDMG <- rbind(NBH20vOCDMRDEG, NBH5vOCDMRDEG)
write.csv(allNBHDMRDMG, "NBH-DMR-DEG-overlaps.csv", quote = FALSE, row.names = FALSE)
```

For each contrast, only one gene with a DMR is present in the DEG data, but these genes are not differentially expressed. For the hypoxia vs. outside control contrast, there are a few more genes that were included in the DEG data set that came up in the `closestBed` analysis, but again, none of these genes are differentially expressed.

# `BAT_correlating`

I [ran an analysis](https://github.com/yaaminiv/killifish-hypoxia-RRBS/blob/main/code/08-BAT-correlating.sh) using [`BAT_correlating`](http://legacy.bioinf.uni-leipzig.de/Software/BAT/dmrs/#bat_correlating). This analysis calculates correlations between methylation in a DMR and expression in an overlapping gene. I will take this output and use it to create figures.

## Normoxia vs. outside control

### Format data

```{r}
NBH20vOCBATcorrelating <- read.delim("BAT_correlating/20_OC_N/correlation_output/20_OC_N_correlation.txt", sep = "\t", header = TRUE, skip = 1, na.strings = c("NaN", "NA")) %>%
  dplyr::select(-1) %>% 
  filter(is.na(rho.p.val) == FALSE) #Import output from BAT_correlating. Skip the first line to allow column names to be recognized. Input specific NA strings. Remove the first column. Keep rows that have correlation p-values
colnames(NBH20vOCBATcorrelating)[1] <- "gene.name" #Rename column with gene ID
NBH20vOCBATcorrelating <- NBH20vOCBATcorrelating %>%
  left_join(NBH20vOCDMR, ., by = "gene.name") %>%
  dplyr::select(-methylation_region) #Join correlating output with DMR dataframe
head(NBH20vOCBATcorrelating) #Confirm import and formatting
```

```{r}
write.table(NBH20vOCBATcorrelating, "BAT_correlating/20_OC_N/correlation_output/20_OC_N_correlation-withAnnot.txt", sep = "\t", quote = FALSE, row.names = FALSE) #Save modified dataframe
```


```{r}
NBH20vOCcorrOnly <- NBH20vOCBATcorrelating %>%
  filter(., is.na(rho.p.val) == FALSE) %>%
  left_join(., (NBH20vOCavgMeth %>% 
                  rename_with(., function(x) paste0(x, sep = "_meth")) %>%
                  rownames_to_column(., var = "DMR")), by = "DMR") %>%
  left_join(., NBH20geneExp, by = "gene.name") %>%
  left_join(., NBHOCgeneExp, by = "gene.name") %>% 
  dplyr::select(-c(chr, DMR.start, DMR.end, direction, gene.start, gene.end, strands, "OC-N3_exp")) %>%
  pivot_longer(cols = "NO_20.N4_meth":"OC_OC.N4_meth", names_to = "sample_meth", values_to = "meth") %>%
  pivot_longer(cols = "20-N1_exp":"OC-N5_exp", names_to = "sample_exp", values_to = "exp") %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "_meth", replacement = "")) %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "NO_20.", replacement = "20-")) %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "OC_OC.", replacement = "OC-")) %>%
  mutate(sample_exp = gsub(x = sample_exp, pattern = "_exp", replacement = "")) %>%
  filter(., sample_meth == sample_exp) %>%
  mutate(sample = sample_meth) %>%
  separate(., col = sample_meth, into = c("treatment", "extra"), sep = "-") %>%
  dplyr::select(-c(sample_exp, extra))
#Take BAT_correlating output and retain rows that have a correlation p-value. Join with average methylation data where the columns are renamed to indicate methylation values. Join with gene expression data. Remove unecessary columns. Pivot longer for methylation and expression data separately. Modify column contents and retain rows where the sample ID for methylation and expression match. Create a new sample ID column. Create a new column for treatment specifications by separating defunct sample_meth column. Remove extra columns
head(NBH20vOCcorrOnly)
```

### Plot

```{r}
plotColors <- c(brewer.pal(9, "GnBu"), 
                brewer.pal(9, "BuGn")) #Create color scheme with greys (outside controls), green-blues, and blue-greens
plotColors <- plotColors[c(9,8,6,18,16,14)] #Pick colors to use in figures. 1 = NB OC, 2 = NB NO, 3 = NB HY
```

```{r}
NBH20vOCcorrText <- data.frame(gene.name = unique(NBH20vOCcorrOnly$gene.name),
                               label = c("rho = 0.39, p = 0.40",
                                         "rho = -0.63, p = 0.13",
                                         "rho = 0, p = 1.00")) #Create a geom with rho and p-value information to label facets
```

```{r}
NBH20vOCcorrOnly %>%
  ggplot(aes(x = meth, y = exp, color = treatment)) +
  geom_point() +
  facet_wrap(~gene.name, scales = "free") +
  labs(x = "Mean Methylation (%)", y = "Expression (CPM)") +
  scale_color_manual(name = "Treatment",
                     values = c(plotColors[2], plotColors[1]),
                     labels = c("Mild", "Control")) +
  geom_smooth(method = lm, se = FALSE, 
              linewidth = 0.75, linetype = 2, color = "black") +
  geom_text(NBH20vOCcorrText, inherit.aes = FALSE, 
            mapping = aes(x = -Inf, y = Inf, label = label),
            hjust = -0.05, vjust = 5) +
  theme_classic(base_size= 15) #Create a scatterplot with methylation and expression data. Color by treatment and facet by gene. Relabel x- and y-axes. Add linear regression line and text labels.
ggsave("figures/20_OC_N-corrPlots.pdf", height = 8.5, width = 11)
```

## Hypoxia vs. outside control

### Format data

```{r}
NBH5vOCBATcorrelating <- read.delim("BAT_correlating/5_OC_N/correlation_output/5_OC_N_correlation.txt", sep = "\t", header = TRUE, skip = 1, na.strings = c("NaN", "NA")) %>%
  dplyr::select(-1) %>% 
  filter(is.na(rho.p.val) == FALSE) #Import output from BAT_correlating. Skip the first line to allow column names to be recognized. Input specific NA strings. Remove the first column. Keep rows that have correlation p-values
colnames(NBH5vOCBATcorrelating)[1] <- "gene.name" #Rename column with gene ID
NBH5vOCBATcorrelating <- NBH5vOCBATcorrelating %>%
  left_join(NBH5vOCDMR, ., by = "gene.name") %>%
  dplyr::select(-methylation_region) #Join correlating output with DMR dataframe
head(NBH5vOCBATcorrelating) #Confirm import and formatting
```

```{r}
write.table(NBH5vOCBATcorrelating, "BAT_correlating/5_OC_N/correlation_output/5_OC_N_correlation-withAnnot.txt", sep = "\t", quote = FALSE, row.names = FALSE) #Save modified dataframe
```

```{r}
NBH5vOCcorrOnly <- NBH5vOCBATcorrelating %>%
  filter(., is.na(rho.p.val) == FALSE) %>%
  left_join(., (NBH5vOCavgMeth %>% 
                  rename_with(., function(x) paste0(x, sep = "_meth")) %>%
                  rownames_to_column(., var = "DMR")), by = "DMR") %>%
  left_join(., NBH5geneExp, by = "gene.name") %>%
  left_join(., NBHOCgeneExp, by = "gene.name") %>% 
  dplyr::select(-c(chr, DMR.start, DMR.end, direction, gene.start, gene.end, strands, "OC-N3_exp")) %>%
  pivot_longer(cols = "HY_5.N1_meth":"OC_OC.N4_meth", names_to = "sample_meth", values_to = "meth") %>%
  pivot_longer(cols = "5-N1_exp":"OC-N5_exp", names_to = "sample_exp", values_to = "exp") %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "_meth", replacement = "")) %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "HY_5.", replacement = "5-")) %>%
  mutate(sample_meth = gsub(x = sample_meth, pattern = "OC_OC.", replacement = "OC-")) %>%
  mutate(sample_exp = gsub(x = sample_exp, pattern = "_exp", replacement = "")) %>%
  filter(., sample_meth == sample_exp) %>%
  mutate(sample = sample_meth) %>%
  separate(., col = sample_meth, into = c("treatment", "extra"), sep = "-") %>%
  dplyr::select(-c(sample_exp, extra))
#Take BAT_correlating output and retain rows that have a correlation p-value. Join with average methylation data where the columns are renamed to indicate methylation values. Join with gene expression data. Remove unecessary columns. Pivot longer for methylation and expression data separately. Modify column contents and retain rows where the sample ID for methylation and expression match. Create a new sample ID column. Create a new column for treatment specifications by separating defunct sample_meth column. Remove extra columns
head(NBH5vOCcorrOnly)
```

### Plot

```{r}
NBH5vOCcorrOnly %>%
  ggplot(aes(x = meth, y = exp, color = treatment)) +
  geom_point() +
  facet_wrap(~gene.name, scales = "free") +
  labs(x = "Mean Methylation (%)", y = "Expression (CPM)") +
  scale_color_manual(name = "Treatment",
                     values = c(plotColors[3], plotColors[1]),
                     labels = c("Severe", "Control")) +
  geom_smooth(method = lm, se = FALSE, 
              linewidth = 0.75, linetype = 2, color = "black") +
  theme_classic() #Create a scatterplot with methylation and expression data. Color by treatment and facet by gene. Relabel x- and y-axes. Add linear regression line and text labels.
ggsave("figures/5_OC_N-corrPlots.pdf", height = 8.5, width = 11)
```

```{r}
NBH5vOCcorrOnly %>%
  filter(., rho.p.val < 0.05) %>%
  dplyr::select(gene.name, rho, rho.p.val, distance) %>%
  unique()
```

```{r}
NBH5vOCcorrText <- data.frame(gene.name = (NBH5vOCcorrOnly %>%
                                             filter(., rho.p.val < 0.05) %>%
                                             dplyr::select(gene.name) %>%
                                             unique()),
                              label = c("rho = 0.86, p = 0.02",
                                        "rho = 0.86, p = 0.01",
                                        "rho = 0.77, p = 0.04")) #Create a geom with rho and p-value information to label facets
```

```{r}
NBH5vOCcorrOnly %>%
  filter(., rho.p.val < 0.05) %>%
  ggplot(aes(x = meth, y = exp, color = treatment)) +
  geom_point() +
  facet_wrap(~gene.name, scales = "free") +
  labs(x = "Mean Methylation (%)", y = "Expression (CPM)") +
  scale_color_manual(name = "Treatment",
                     values = c(plotColors[3], plotColors[1]),
                     labels = c("Severe", "Control")) +
  geom_smooth(method = lm, se = FALSE, 
              linewidth = 0.75, linetype = 2, color = "black") +
  geom_text(NBH5vOCcorrText, inherit.aes = FALSE,
            mapping = aes(x = -Inf, y = Inf, label = label),
            hjust = -0.05, vjust = 5) +
  theme_classic(base_size = 15) #Create a scatterplot with methylation and expression data. Color by treatment and facet by gene. Relabel x- and y-axes. Add linear regression line and text labels.
ggsave("figures/5_OC_N-corrPlots-sigCorr.pdf", height = 8.5, width = 11)
```
