---
title: "08-annotate-DMR"
author: "Yaamini Venkataraman"
date: "2024-08-07"
output: html_document
---

In this script I will examine overlaps between DMR and DEG for specific contrasts, then visualize DMR patterns. As a reminder, there were no DMR found for Scorton Creek populations, so this script will focus only on New Bedford Harbor populations when examining DMR-DEG overlaps.

# Set up R Markdown document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/yaaminivenkataraman/Documents/killifish-hypoxia-RRBS/output/08-annotate-DMR/new-genome/") #Set root directory
```

```{r}
getwd()
```

#Install packages

```{r}
#install.packages("tidyverse")
#install.packages("RColorBrewer")
require(tidyverse)
require(RColorBrewer)
```

```{r}
sessionInfo()
```

# Import and format data

## DMR data

```{r}
NBH20v0CDMR <- read.delim("20_OC_N_DMR.closestGene.bed", header = FALSE, 
                          col.names = c("chr", "DMR.start", "DMR.end", "DMR", "direction", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strands", "V13", "V14", "distance")) %>%
  separate_wider_delim(., cols = "V14", delim = ";Dbxref", names = c("gene.name", "trash")) %>%
  mutate(., gene.name = gsub(x = gene.name, pattern = "ID=gene-", replacement = "")) %>%
  dplyr::select(chr, DMR.start, DMR.end, DMR, direction, gene.name, gene.start, gene.end, strands, distance)
  #Import DMR data and add column names. Separate column with gene name by a specific delimiter, then remove characters before gene name. Retain only necessary columns. 
head(NBH20v0CDMR) #Confirm formatting
```

```{r}
NBH5vOCDMR <- read.delim("5_OC_N_DMR.closestGene.bed", header = FALSE, 
                          col.names = c("chr", "DMR.start", "DMR.end", "DMR", "direction", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strands", "V13", "V14", "distance")) %>%
  separate_wider_delim(., cols = "V14", delim = ";Dbxref", names = c("gene.name", "trash")) %>%
  mutate(., gene.name = gsub(x = gene.name, pattern = "ID=gene-", replacement = "")) %>%
  dplyr::select(chr, DMR.start, DMR.end, DMR, direction, gene.name, gene.start, gene.end, strands, distance)
  #Import DMR data and add column names. Separate column with gene name by a specific delimiter, then remove characters before gene name. Retain only necessary columns. 
head(NBH5vOCDMR) #Confirm formatting
```

```{r}
NBH20v0Cmetilene <- read.delim("../../05-analysis/new-genome/summarize/missing_1/20_OC_N/N_metilene_NO_OC.txt", header = TRUE) #Import metilene data used to identify DMR. This dataset has the methylation value for each CpG in the DMR.
head(NBH20v0Cmetilene) #Confirm import
```

```{r}
NBH20v0Cpositions <- data.frame() #Create empty dataframe
for (i in 1:nrow(NBH20v0CDMR)) {
  temp <- NBH20v0Cmetilene %>%
  filter(., pos >= NBH20v0CDMR$DMR.start[i] & pos <= NBH20v0CDMR$DMR.end[i]) %>%
  mutate(., DMR = NBH20v0CDMR$DMR[i])
  NBH20v0Cpositions <- rbind(NBH20v0Cpositions, temp)
} #Use a for loop to filter CpG data for each DMR. Filter positions based on start and end positions of each DMR. Add a column for DMR identity. Save as a temporary dataframe. Use rbind to append information to empty dataframe
colnames(NBH20v0Cpositions) <- c(colnames(NBH20v0Cmetilene), "DMR") #Assign column names based on previous dataframe and add name for the DMR column
head(NBH20v0Cpositions) #Confirm formatting
```


```{r}
NBH20v0CDMR
```


## DEG data

```{r}
NBH20vOCDEG <- read.csv("../../../data/NEW-n.NBHvso.NBH.csv") #Import DEG data
colnames(NBH20vOCDEG)[1] <- "gene.name"
head(NBH20vOCDEG)
```

```{r}
NBH5vOCDEG <- read.csv("../../../data/NEW-h.NBHvso.NBH.csv") #Import DEG data
colnames(NBH5vOCDEG)[1] <- "gene.name"
head(NBH5vOCDEG)
```

# Heatmaps of DMR

I will make heatmaps showing methylation level of DMR across samples for the normoxia vs. outside control and hypoxia vs. outside control comparisons.

```{r}
head(NBH20v0CDMR)
```

```{r}

```


# Overlaps between DEG and DMR

```{r}
NBH20v0CDMRDEG <- inner_join(NBH20v0CDMR, NBH20vOCDEG, by = "gene.name") %>%
  mutate(contrast = "20_OC_N") #Use inner join to identify DMR that overlap with DEG.
NBH20v0CDMRDEG
```

```{r}
NBH5v0CDMRDEG <- inner_join(NBH5vOCDMR, NBH5vOCDEG, by = "gene.name") %>%
  mutate(contrast = "5_OC_N") #Use inner join to identify DMR that overlap with DEG.
NBH5v0CDMRDEG
```

```{r}
allNBHDMRDMG <- rbind(NBH20v0CDMRDEG, NBH5v0CDMRDEG)
write.csv(allNBHDMRDMG, "NBH-DMR-DEG-overlaps.csv", quote = FALSE, row.names = FALSE)
```

For each contrast, only one gene with a DMR is present in the DEG data, but these genes are not differentially expressed. For the hypoxia vs. outside control contrast, there are a few more genes that were included in the DEG data set that came up in the `closestBed` analysis, but again, none of these genes are differentially expressed.

# DMR Figure

